oolong: An R package for validating  automated content analysis tools

### 2020-07-21 paper/paper.rmd
port Marius's idea back to paper.rmd, ref #31

# Summary

Oolong is an R package providing functions for validating common automated content analysis techniques such as topic modeling and dictionary-based methods. This package is designed for R users needing to validate these methods. Typical users of oolong are communication and political scientists, among others.

@@

Validity is a requirement of content analysis [@krippendorff2018content; @neuendorf2016content]. Validation of automated methods has been called for by many scholars, e.g. @grimmer2013text; @ribeiro2016sentibench; @van2018communication.

Oolong makes it easy to generate standard validation tests suggested by @chang2009reading and @song2020validations.

@@

The paper by @dimaggio2013exploiting conceptualizes validation of automated methods as three different operations and the three operations supplement each other. These three operations are: 1) *statistical* validation -- to see if the model results agree with the assumptions of the model. Examples of statistical validation are calculation of pointwise mutual information, perplexity or semantic coherence of a topic model. 2) *semantic* validation -- to see if the model results are semantically meaningful. This procedure involves comparing model results with human judgment [@grimmer2011general]. 3) *predictive* validation -- to see if the model results can predict external events [@quinn2010analyze]. For example, one can study whether external events can explain surges in attention to a topic extracted by a topic model.

This package focuses on semantic validation. The reason is threefold. First, there is existing architecture for conducting statistical validation and predictive validation. Topic modeling packages such as `text2vec` [@selivanov2020tex2vec] and `topicmodels` [@bettina2011topicmodels] provide functions to calculate metrics such as perplexity. Packages such as `textmineR` [@jones2019textminer], `stminsights` [@schwemmer2018stminsights] and `LDAvis` [@sievert2015ldavis] offers additional methods for statistical validation and predictive validation. As of writing, `tosca` [@koppers2020tosca] is the only package dealing with semantic validation. But the text-based interface might pose challenges to human annotators and it can only support topic models from one package [@change2015lda].

Second, results from statistical validation do not always agree with those from semantic validation. For example, a topic model with a lower perplexity does not have a better interpretability [@chang2009reading]. Of course, there are also metrics from statistical validation that are shown to be correlated with semantic validity, e.g. semantic coherence [@mimno2011optimizing]. Calculation of semantic coherence is recommended in the best practice paper by @maier2018applying. Nonetheless, conducting only statistical validation is not adequate because 3 validation operations supplement each other.

Finally, predictive validation is dependent on research questions and thus it is difficult to be generalized as a reusable software framework. Additionally, the relationship between external (sociopolitical) events and the results from automated content analysis tools is usually what social scientists are eager to study, c.f. using topic models for information retrieval [@yi2008evaluating]. We do not believe social scientists would forget to conduct any form of predictive validation for their topic models.

Oolong focuses on semantic validation and the "human-in-the-loop" procedure for it has been standardized [@chang2009reading; @song2020validations]. The procedure proposed by @chang2009reading has been adopted in subsequent social science studies as the gold standard to validate topic models, e.g. @bohr2020reporting, @chuang2015topiccheck, and @miller2017australia.

# Validating topic models

@@

# Validating dictionary-based methods

>>>>

@@

Oolong is an R package providing functions for semantic validation of topic modeling and dictionary-based methods, two main tools for doing automated content analysis [@boumans2016taking;@gunther2016word].

While the validation of statistical properties of topics models is well established, the substantive meaning of categories uncovered is often less clear and their interpretation reliant on "intuition" or "eyeballing". As @chang2009reading [p. 1] put it: "qualitative evaluation of the latent space" or figuratively, reading tea leaves.

The story for dictionary-based methods is not better. Researchers usually assume these dictionaries have built-in validity and use them directly in their research. However, multiple validation studies [@boukes2020whatsthetone;@gonzalez2015signals;@ribeiro2016sentibench] demonstrate these dictionaries have very limited criterion validity.

Oolong provides a set of tools to objectively judge substantive interpretability to applied users in disciplines such as political science and communication science. It allows standardized content based testing of topic models as well as dictionary-based methods with clear numeric indicators of semantic validity. Oolong makes it easy to generate standard validation tests suggested by @chang2009reading and @song2020validations.

@@

Validity is a requirement of content analysis [@krippendorff2018content; @neuendorf2016content]. Validation of automated methods has been called for by many scholars, e.g. @grimmer2013text; @ribeiro2016sentibench; @van2018communication. The paper by @dimaggio2013exploiting conceptualizes validation of automated methods as three different operations and the three operations supplement each other. These three operations are: 1) *statistical* validation -- to see if the model results agree with the assumptions of the model. Examples of statistical validation are calculation of pointwise mutual information, perplexity or semantic coherence of a topic model. 2) *semantic* validation -- to see if the model results are semantically meaningful. This procedure involves comparing model results with human judgment [@grimmer2011general]. 3) *predictive* validation -- to see if the model results can predict external events [@quinn2010analyze]. For example, one can study whether external events can explain surges in attention to a topic extracted by a topic model.

This package focuses on semantic validation for three reasons: 
First, there is existing architecture for conducting statistical validation and predictive validation. Topic modeling packages such as `text2vec` [@selivanov2020tex2vec], `topicmodels` [@bettina2011topicmodels], and `textmineR` [@jones2019textminer] provide functions to calculate metrics such as perplexity and semantic coherence. Packages such as `stminsights` [@schwemmer2018stminsights] and `LDAvis` [@sievert2015ldavis] offer additional qualitative methods for predictive validation. As of writing, `tosca` [@koppers2020tosca] is the only package dealing with semantic validation. But the text-based interface might pose challenges to human annotators and it can only support topic models from the `lda` package [@change2015lda].

Second, results from statistical validation do not always agree with those from semantic validation. For example, a topic model with a lower perplexity does not have a better interpretability [@chang2009reading]. Of course, there are also metrics from statistical validation that are shown to be correlated with semantic validity, e.g. semantic coherence [@mimno2011optimizing]. Calculation of semantic coherence is recommended in the best practice paper by @maier2018applying. Nonetheless, conducting only statistical validation is not adequate because these three validation operations supplement each other.

Finally, predictive validation is dependent on research questions and thus it is difficult to be generalized as a reusable software framework. Additionally, the relationship between external (sociopolitical) events and the results from automated content analysis tools is usually what social scientists are eager to study, c.f. using topic models for information retrieval [@yi2008evaluating]. We do not believe social scientists would ignore conducting any form of predictive validation.

Oolong focuses on semantic validation. The package provides the "human-in-the-loop" semantic validation procedures suggested by @chang2009reading and @song2020validations. The procedure proposed by @chang2009reading has been adopted in subsequent social science studies as the gold standard to validate topic models, e.g. @bohr2020reporting, @chuang2015topiccheck, and @miller2017australia.

# Semantic validation of topic models

@@

# Semantic validation of dictionary-based methods
